---
title: "Assignment 3B Window Functions"
format: pdf
editor: visual
---

## Introduction

This assignment will test my use of window functions, which unlike aggregate functions, make an output that is available for each row of data. I will use functions in dplyr to calculate year-to-date and a six-day moving averages. Although I am not using window functions in SQL, I would still like to use SQL in some capacity, even if it just simply creating a table like last week's assignment.

## Challenges

While this assignment is fairly straightforward, I do anticipate some issues with using window functions for the first time. Like everything I have created for this class, it will take time for me to grasp the when window functions are called for and how to use them most effectively.

```{r}
library(RKaggle)
library(tidyverse)
library(slider)
```

```{r}
temperatures <- RKaggle::get_dataset("sudalairajkumar/daily-temperature-of-major-cities")

```

Let's start by filtering this data set down. Since the data run from 1/1/1995 to 5/13/2020, I chose 2020 to be the year to look at. I chose to look simply at New York City.

```{r}
temperatures <- temperatures |> 
  filter(City == "New York City",
         Year == "2020")
```

First, I'll find the year-to-date average. This can be computed by taking the mean of all daily average temperatures.

```{r}
temperatures |> 
  summarise(ytd_temp = mean(AvgTemperature))
```

This shows the average daily temperature from 1/1/2020 - 5/13/2020 to be a brisk 45 degrees.

Now I'll use the slide_mean function from the slider package to calculate the 6-day rolling average. Note, the "before" argument takes in the number of observations to fit into the window *before* the observation in the row in question. This makes sense when looking at a moving average with a "window" of size k. $$\frac{1}{k}\sum_{i = n - k + 1}^{n} x_{i}$$

```{r}
temperatures <- temperatures |> 
  mutate(six_day_avg = round(
    slide_mean(AvgTemperature, before = 5), digits = 1))

temperatures$six_day_avg


```
